---
title: 'Econometrics Puzzler #4: Rescaling the Reduced Form'
author: Francis J. DiTraglia
date: '2026-08-14'
subject: 'Inference & uncertainty'
eyebrow: 'Puzzler No. 4'
categories:
- econometrics
- causal inference
- puzzler
tags:
- instrumental variables
- asymptotics
---

Consider the simplest textbook instrumental variables (IV) model with one endogenous regressor $X$ and a single valid instrument $Z$. You are given the estimated slope coefficient $\hat{\gamma}$ from the reduced form regression of $Y$ on $Z$ along with its standard error $\text{SE}(\hat{\gamma})$. You are also given the estimated slope coefficient $\hat{\pi}$ from the first stage regression of $X$ on $Z$. The IV estimate equals  $\hat{\gamma}/ \hat{\pi}$. Does its standard error equal  $\text{SE}(\hat{\gamma})/\hat{\pi}$?

## Solution

::: {.callout-note .solution icon=false collapse="true" appearance="simple" title="Solution"}

Here's why it's tempting to answer *yes*. If $Y$ is measured in dollars and $Z$ is measured in grams then $\hat{\gamma}$ and $\text{SE}(\hat{\gamma})$ are both measured in dollars per gram. Now suppose that you wanted to express both quantities in dollars per *kilogram* rather than dollars per gram. No problem: simply multiply each by 1000. The IV logic from above *looks* analogous: the IV estimator is simply a rescaled version of the reduced form. So does the same relationship hold for the standard error? Well there's an equally tempting reason to say *no*. Since $\hat{\pi}$ is estimated, rather than a known constant, dividing by it should *add* sampling variability of its own. This reasoning suggests that $\text{SE}(\hat{\gamma})/\hat{\pi}$ should be *too small*. So which is it? When in doubt, simulate:
```{r, message = FALSE}
set.seed(1848)
n <- 1000
z <- runif(n)
S <- matrix(c(1, 0.5,
                  0.5, 1), 2, 2, byrow = TRUE)

library(mvtnorm)
errors <- rmvnorm(n, mean = c(0, 0), sigma = S)
u <- errors[, 1]
v <- errors[, 2]

pi <- 0.7
beta <- 1.2

x <- pi * z + v
y <- beta * x + u

library(ivreg)
library(tidyverse)
library(broom)

get_coef_se <- function(results, varname) {
  results |>
    tidy() |>
    filter(term == varname) |>
    select(estimate, std.error)
}

iv <- ivreg(y ~ x | z) |>
  get_coef_se('x')

first_stage <- lm(x ~ z) |>
  get_coef_se('z')

reduced_form <- lm(y ~ z) |>
  get_coef_se('z')
```
The IV slope estimate is exactly equal to the ratio of reduced form and first stage slopes, as it should be
```{r}
c(est_by_hand = reduced_form$estimate / first_stage$estimate,
  est_ivreg = iv$estimate)
```
but the IV *standard error* does not share the same proportionality relationship to the reduced form standard error:
```{r}
c(se_by_hand = reduced_form$std.error / first_stage$estimate,
  se_ivreg = iv$std.error)
```

So the answer is *no*, but it seems that *both* of the tempting intuitions have failed us here: the naive standard error isn't too small, it's nearly *twice* as large as the correct one. What's going on?

To answer this question, we'll start by writing down the structural equation $Y = \beta X + U$ and the first-stage $X = \pi Z + V$.^[To keep the math simple, I implicitly assume here that $(Y,X,Z)$ are mean zero so that there is no need to include an intercept in either equation. Equivalently, we could simply subtract the mean from each random variable before proceeding.] Combining them gives the reduced form:
$$
\begin{aligned}
Y &= \beta X + U = \beta(\pi Z + V) + U\\
&= (\beta\pi) Z  + (\beta V + U)\\
&= \gamma Z + \varepsilon
\end{aligned}
$$
where $\gamma \equiv \beta \pi$ and $\varepsilon \equiv \beta V + U$.
Now, the IV estimator can be expanded according to
$$
\begin{aligned}
\hat{\beta} &= (\underline{Z}'\underline{X})^{-1}(\underline{Z}'\underline{Y}) = (\underline{Z}'\underline{X})^{-1}\left[\underline{Z}'(\beta \underline{X} + \underline{U})\right]\\
&= \beta + (\underline{Z}'\underline{X})^{-1}(\underline{Z}'\underline{U})
\end{aligned}
$$
where $\underline{Z}'$ denotes the row vector $(Z_1, ... Z_n)$ and analogously for $\underline{Y}, \underline{X}$ and $\underline{U}$. Thus, subtracting $\beta$ from both sides and multiplying by $\sqrt{n}$,
$$
\sqrt{n}(\hat{\beta} - \beta) = \left(\frac{\underline{Z}'\underline{X}}{n}\right)^{-1}\left(\frac{\underline{Z}'\underline{U}}{\sqrt{n}}\right) \xrightarrow{d} \frac{1}{\mathbb{E}(ZX)}  \times N\left(0, \mathbb{E}[Z^2U^2]\right)
$$
by the central limit theorem. Proceeding similarly for the reduced form estimator:
$$
\begin{aligned}
\hat{\gamma} &= (\underline{Z}'\underline{Z})^{-1}(\underline{Z}'\underline{Y}) = (\underline{Z}'\underline{Z})^{-1}\left[\underline{Z}'(\gamma \underline{Z} + \underline{\varepsilon})\right]\\
&= \gamma + (\underline{Z}'\underline{Z})^{-1}(\underline{Z}'\underline{\varepsilon})
\end{aligned}
$$
and therefore,
$$
\begin{aligned}
\sqrt{n}(\hat{\gamma} - \gamma) = \left( \frac{\underline{Z}'\underline{Z}}{n}\right)^{-1}\left( \frac{\underline{Z}'\underline{\varepsilon}}{\sqrt{n}}\right) \xrightarrow{d} \frac{1}{\mathbb{E}(Z^2)} \times N(0, \mathbb{E}[Z^2\varepsilon^2])
\end{aligned}
$$
To keep the math simple I'll assume that $Z$ is *independent* of both $U$ and $V$, as in the simulation above.^[For more on the various notions of "independence" in econometrics see [this post](/post/why-econometrics-is-confusing-part-ii-the-independence-zoo/).] Under this assumption $Z$ is *also* independent of $\varepsilon$ and thus
$$
\begin{aligned}
\mathbb{E}(Z^2 \varepsilon^2) &= \mathbb{E}(Z^2)\mathbb{E}(\varepsilon^2) = \sigma_z^2 \sigma_\varepsilon^2\\
\mathbb{E}(Z^2 U^2) &= \mathbb{E}(Z^2)\mathbb{E}(U^2) = \sigma_z^2 \sigma_u^2
\end{aligned}
$$
since $Z, \varepsilon$ and $U$ are mean zero. Thus, the asymptotic variances of the IV and reduced form estimators are given by
$$
\begin{aligned}
\text{AVAR}(\widehat{\beta}) &= \left(\frac{1}{\mathbb{E}[ZX]}\right)^2 \mathbb{E}(Z^2 U^2) = \left(\frac{1}{\pi \sigma_z^2}\right)^2 \sigma_z^2 \sigma_u^2 = \frac{\sigma_u^2}{\pi^2 \sigma_z^2}\\
\text{AVAR}(\widehat{\gamma}) &= \left(\frac{1}{\mathbb{E}[Z^2]}\right)^2 \mathbb{E}(Z^2 \varepsilon^2) = \left( \frac{1}{\sigma_z^2}\right)^2 \sigma_z^2 \sigma_\varepsilon^2 = \frac{\sigma_\varepsilon^2}{\sigma_z^2}
\end{aligned}
$$
since the first-stage slope coefficient $\pi$ equals  $\mathbb{E}[ZX]/ \text{Var}(Z)$. And now we have our answer! Taking square roots gives
$$
\sqrt{\text{AVAR}(\hat{\beta})} = \frac{\sigma_u}{\pi \sigma_z}, \quad
\sqrt{\text{AVAR}(\hat{\gamma})} = \frac{\sigma_\varepsilon}{\sigma_z}
$$
The respective standard errors are *estimators* of these quantities, divided by $\sqrt{n}$. Fundamentally, the standard error of the IV estimator is governed by the variation in the *structural error* $U$ while that of the reduced form estimator is governed by the variation in the *reduced form error*. It follows that
$$
\sqrt{\text{AVAR}(\hat{\beta})} = \left(\frac{\sigma_u}{\sigma_\varepsilon}\right) \cdot \frac{\sqrt{\text{AVAR}(\hat{\gamma})}}{\pi}
$$
so there **really is** a factor of $\pi$ in the relationship between the IV and reduced form standard errors! The trouble is that there is a second factor,  $\sigma_u/\sigma_\varepsilon$, and this *does not* equal one. Instead,
$$
\sigma_\varepsilon^2 = \text{Var}(\beta V + U) = \beta^2 \sigma_v^2 + \sigma_u^2 + 2\beta \,\text{Cov}(U,V)
$$
so that $\sigma_\varepsilon^2 - \sigma_u^2 = \beta^2 \sigma_v^2 + 2\beta \text{Cov}(U,V)$. The first term in the difference, $\beta^2 \sigma_v^2$, is always positive but the second is negative whenever the sign of $\beta$ disagrees with that of $\text{Cov}(U,V)$. This means that $\sqrt{\text{AVAR}(\hat{\beta})}$ can either be larger or smaller than $\sqrt{\text{AVAR}(\hat{\gamma})}/\pi$ depending on the values of the variances and covariances of $(U,V)$. In our simulation $\beta$ and $\text{Cov}(U,V)$ are both positive, so $\sigma_\varepsilon > \sigma_u$. This is why the naive standard error was too large.

What about our second intuition? Writing the reduced form estimate in terms of the first stage estimate gives
$$
\hat{\gamma} = \beta \hat{\pi} + (\underline{Z}'\underline{Z})^{-1}(\underline{Z}'\underline{U}).
$$
This relationship is exact: it holds in *every sample*. Subtracting $\gamma = \beta \pi$ from both sides gives $\hat{\gamma} - \gamma = \beta(\hat{\pi} - \pi) + (\underline{Z}'\underline{Z})^{-1}(\underline{Z}'\underline{U})$. Part of the sampling error in $\hat{\gamma}$ is simply $\beta$ times the sampling error in $\hat{\pi}$. What remains is the term involving $\underline{Z}'\underline{U}$. Dividing by $\hat{\pi}$ turns the shared component into $\beta$, a constant. This moves the first stage error out of the numerator and into the denominator. We are left with $\hat{\beta} = \beta + \underline{Z}'\underline{U}/\underline{Z}'\underline{X}$. The denominator is still random, but when we rescale it by $n$ and apply the central limit theorem, $\underline{Z}'\underline{X}/n$ converges in probability to $\mathbb{E}(ZX)$, a constant. Its reciprocal is precisely the $1/\mathbb{E}(ZX)$ multiplying the normal limit in our expression for $\sqrt{n}(\hat{\beta} - \beta)$ above.

Finally, we can check to see that our calculations are correct by comparing them to the simulation results from above. The standard errors reported by `lm` and `ivreg` are finite sample counterparts of our asymptotic standard errors, so the same relationship should hold with $\hat{\sigma}_u$ and $\hat{\sigma}_\varepsilon$ in place of $\sigma_u$ and $\sigma_\varepsilon$. The reduced form residuals $\hat{\varepsilon}_i$ and the first stage residuals $\hat{V}_i$ give us the structural residuals, since $\hat{\varepsilon}_i - \hat{\beta}\hat{V}_i = Y_i - \hat{\beta}X_i = \hat{U}_i$:
```{r}
beta_hat <- reduced_form$estimate / first_stage$estimate
eps_hat <- residuals(lm(y ~ z))
u_hat <- eps_hat - beta_hat * residuals(lm(x ~ z))

c(se_by_hand = (sd(u_hat) / sd(eps_hat)) *
    reduced_form$std.error / first_stage$estimate,
  se_ivreg = iv$std.error)
```

:::
